\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2017-12-07 document/XSGRB_sample.tex
Adding RRM text

\abstract{
In this work we present the spectra of \textit{all} $\gamma$-ray bursts (GRBs)
afterglows that have \textit{ever} been observed at the VLT/X-shooter up to the end of 2016. This
totals 137 spectroscopic observation of 118 individual bursts, of which
redshifts have been secured for 91 \% covering a redshift range from 0.059 to
$\sim$ 8.5. Based on a set of unbiasing selection criteria, the follow-up effort
has been guided by the \textit{Swift}-telescope to produce a homogeneous sample
of afterglows for 86 spectra, representing the underlying \textit{Swift} GRB population.
We here provide a public release of all the spectra, including continuum
estimates and telluric corrections. We provide an assessment of the degree of
completeness with respect to the parent GRB population, in terms of the X-ray
properties of the bursts in the sample and find that the sample presented here
can not be significantly distinguished from the \textit{Swift} sample.
For all bursts (39) for which it is possible, we provide a measurement of the
optically derived hydrogen column density, increasing the total number of
published HI column densities by $\sim$ 50 \% and characterize the distribution
compared with literature values.
We constrain the fraction of dark burst to be 27\% - 42\% \todo{Preliminary} and tentatively confirm previous
results that larger optical darkness leads to increased X-ray absorption.
%This dataset provides a unique possibility to study the ISM across cosmic time.
}

\keywords{Gamma-ray burst: individual: GRB~120815A --- galaxies: high-redshift --- ISM: molecules --- dust, extinction }

@@

Gamma-Ray bursts (GRBs) are interesting as they provide the opportunity to
derive new insights into a very wide range of topics in astrophysics. Examples
range from small scale phenomena relating to the magnetars, properties of highly
relativistic jets, hyper/supernova explosions, the interstellar medium, dust extinction
curves, starburst galaxies, chemical and molecular abundances, escape of ionizing 
radiation, the ionization state of the intergalactic medium, intervening
absoption systems to standard candles in cosmology.  In that sense GRB research
is very multidisciplinary.

The {\it Swift} satellite \citep{Gehrels2009}, which was launched in 2004, has
made it possible to harvest much of the great potential in using GRBs as probes
of galaxies in the intergalactic medium, which was already hinted at by results
from earlier missions \citep[e.g.][]{Ricker2004}.  A crucial aspect of the great
success of the {\it Swift} mission has been the extensive ground based
follow-up observations of the afterglows and of the host galaxies of the GRBs.
This work has involved a large community of researchers both inside and outside
the official {\it Swift} team. This fruitful collaboration has been greatly
facilitated by the open data access policy of the {\it Swift} mission.

In the beginning of the "{\it Swift} era" most of the follow-up spectroscopy
was secured using low-resolution spectrographs 
\citep[typically
R=$\lambda/\Delta\lambda$$<$1000, e.g.][]{Fynbo2009}. Low-resolution
spectroscopy is powerful for a number of reasons, first of all is the ability
to secure information for even very faint targets. This allows the measurements
of a number of important parameters such as redshifts, spectral slopes,
and extinction. For
a handful of very bright afterglows high-resolution (typically R$>$20000)
spectra were secured and for these events much more information about
conditions inside the host galaxies were extracted 
\citep[e.g.,][]{Fiore2005, Thone2007, Prochaska2007, Vreeswijk2007, Castro-Tirado2010}.

The X-shooter spectrograph \citep{Vernet2011} is the first of the 2nd
generation instruments at the ESO Very Large Telescope (VLT). It is a
high-throughput echelle spectrograph covering the full spectral range from the
UV atmospheric cutoff around 300 nm to the near-infrared (near-IR) K-band 
ending around 2500 nm with three spectral
arms (UVB, VIS, NIR). It was designed very much with transient source follow-up
in mind as the fading luminosities of such sources makes in urgent to secure as
extensive coverage as possible in the shortest possible time. At the same time
the resolution was set to be in the range 4000--9000 in order to be able to get
a large useful spectral range between the many sky-background emission lines in
the red and near-IR spectral ranges. In this paper we present the results
of a dedicated effort over the years 2009 -- 2016 to use the X-shooter
spectrograph to secure spectroscopic observations of
afterglows and host galaxies of GRBs detected by {\it Swift}. 
All data resulting from the survey are publicly availble in reduced form.

The paper is organized in the following way: In Sect.~\ref{sample} we describe
the sample including the sample selection criteria and the observational
strategy.  In Sec.t~\ref{results} we describe the results of the survey, i.e.
the efficiency of the follow-up effort and the characteristics of the observed
bursts. We also here assess the completeness of the finally realized sample.
Finally, we offer our conclusions in Sect.~\ref{conclusions}.

@@

Being of transient nature, it is difficult to impose strong sample selection
criteria on GRBs, without hampering the follow-up effort. Many natural follow-up
restrictions exists already, being it weather conditions, pointing restrictions
of the telescope or poorly localized bursts as reported by the alerting
telescope. To maximize the return of the follow-up campaign we have chosen a few
selection criteria that attempts to provide an unbiased selection of bursts,
while allowing for a high follow-up success rate. The importance of defining
unbiasing selection criteria has been highlighted previously
\citep{Salvaterra2012, Hjorth2012, Vergani2015, Perley2016a}, when trying to
address the specific underlying distribution functions such as the redshift
distribution, host metallicity distribution, or brightness distribution. When
investigating a specific distribution function, a high degree of
\textit{completeness} is desired meaning that in terms of the parameter of
interest, the largest possible fraction of targets in the sample has
determinations of that parameter \citep{Perley2016b}.

We simultaneously want to minimize any biases against astrophysical conditions
while at the same time maximizing likelihood of observations allowing us to
obtain a higher degree of observing completeness. By restricting the selection
criteria to conditions local to Galaxy and therefore irrespective of intrinsic
GRB properties, the hope is that the sample collect represents the underlying
distribution of GRBs in a fair way while at the same time, additional unbiasing
cuts can be imposed that can improve the follow-up success. The selection
criteria used here is based on previous successes \cite{Jakobsson2006b,
	Fynbo2009, Hjorth2012}, which has shown to represent the underlying distribution
to a high degree. We test the degree of completeness in Sect.
\ref{completeness}.

The selection criteria are defined as follows:

\begin{enumerate}
	\item GRB triggered onboard by Swift.
	\item XRT started observing within 10 minutes since the GRB; an XRT position must be distributed within 12 hr.
	\item The target must be visible for at least 60 minutes at least 30 degrees above the horizon, with the Sun below -12.
	\footnote{Note that in the P84 proposal the criteria have
	been stated a bit differently, the visibility constraint being replaced by a
	declination + Sun angle constraint. The below criteria are however those
	defining the sample.}
	\item No bright closeby stars.
	\item Galactic $A_V \lesssim 0.5$ mag.
\end{enumerate}

Because our ability to observe GRB afterglows is strongly correlated with the
rapidness and precision of the target positions delivered by the triggering
facilities, by selecting only bursts that have been triggered on board the
\textit{Swift} space telescope \cite{Gehrels2004}, by the Burst Alert Telescope
(BAT) \citep{Barthelmy2005}, we start out with a homogeneously selected sample
where burst characteristics are delivered immediately, allowing for an informed
follow-up strategy. Despite the complexity of the triggering mechanism on board
\textit{Swift} \citep{Band2006}, attempts at inferring properties of the
underlying cosmic GRB population based on the detection thresholds and
triggering algorithms have been made \citep{Lien2014, Graff2016} and thereby by
restricting the follow-up effort to bursts detected by \textit{Swift}, the
limitations of the parent sample is well studied.

Because the localization accuracy of BAT is 1 - 4 arcminutes
\citep{Barthelmy2005}, we additionally require an X-ray position from the X-ray
Telescope (XRT) \citep{Burrows2005} to be distributed to the GCN network
\citep{Barthelmy2000} with 10 hours and to account for observing constraints on
\textit{Swift} that XRT began observations within 10 minutes. The additional
rapidness requirement of the XRT follow-up means that the number of detected
X-ray afterglows for BAT triggers is 100 percent \citep{Burrows2007}. The
\textit{Swift} constraints on the GRB follow-up means that the sample presented
here inherits the \textit{Swift} sample properties, except the additional XRT
constraint. Because the XRT completeness is very high\citep{Burrows2007}, this
should change the parent sample significantly. This issue is addressed again in
Sect. \ref{completeness}.

To allow for observability we additionally require that the GRB is visible
(duh!) from the telescope site at Cerro Paranal, Antofagasta, Región de
Antofagasta, Chile, for a least 1 hour allowing time for the spectroscopic
observations to completed. Since the GRB population is isotropically distributed
seen from Earth and that GRB properties does not depend on position on the sky
\citep{Briggs1996}, this cut does not influence our ability to sample the
underlying GRB population. The same arguments applies to the requirement that
there are no nearby bright stars, which is additionally supplemented by the
requirement what the galactic dust content does not exert an extinction higher
than $A_V \lesssim 0.5$ mag based on the extinction maps by
\citet{Schlegel1998}. Because of  isotropy, these additional cuts should not
influence the optical properties of the bursts themselves, only our ability to
successfully secure observations that allows us to investigate the spectroscopic
properties of GRBs.

We address the succesfullness of these criteria and the actual follow-up effort
in representing the full \textit{Swift} sample ion Sect. \ref{completeness}.

@@

\todo{Will be written by Daniele?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{RRM observations} \label{RRM}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The rapid-response mode is

@@

Observations in this sample has been carried out using the cross-dispersed
echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on two of the four
Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal) during the duration
of this follow-up campaign, which covers the entire lifetime of X-shooter.
Observations have been carried out from the ESO period P84 through P98 under the
following programme IDs: 098.A-0055, 097.A-0036, 096.A-0079, 095.A-0045,
094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088, 089.A-0067,
088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260. A few additional
bursts, not a part of the \textit{statistical sample}, have been included from
092.D-0056(A) (PI: Rau), 092.D-0633(E) (PI: Greiner), and Levan under
095.B-0811(B) (PI: Levan). This sample represents \textit{all} GRBs that have
ever been followed up by X-shooter.

The first GRB followed up is GRB090313, observed 15th of March 2009, which was
while X-shooter was mounted on UT3 during the commissioning of the instrument.
The bursts observed during the commissioning period is not a part of our
statistical sample, but are nonetheless published as part of the bursts observed
by the X-shooter GRB team. The first burst observed after science verification
was completed when X-shooter was moved to UT2 is GRB091018 which thereby
constitute the first burst entering our statistical sample. For all bursts that
fulfill our sample selection criteria, described in Sect. \ref{samplecrit},
spectroscopic follow-up has been attempted with X-shooter. Various conditions
can affect our ability to follow up a given burst, and a discussion of this
effect is included in Sect. \ref{badbursts}.

X-shooter can cover the spectral wavelength region from 300 nm to 2480 nm in a
single exposure, by splitting the light in three separate spectroscopic arms
using dichroics.  This way each arm work as a separate instrument, each
functioning as its own echelle spectrograph.  The ultraviolet blue(UVB) arm
covering 3000 - 550 nm, the visual(VIS) arm covering 550 - 1020 nm, and the
near-infrared()NIR) arm covering 10200 - 2480 nm with the possibility of
applying a k-band blocking filter cutting the coverage of the NIR arm at 1020 -
2100 nm. For all observations, a nod-and-shuffle observing scheme has been
employed, with a nodding throw of 5 \arcsec. The individual observations has
been carried out in a standard ABBA pattern, which allows flexibility in the sky
subtraction method for the different arms.

For the large majority of the bursts we have observed with a slit width of
1\farc0, 0\farc9, and 0\farc9 for the UVB, VIS, and NIR-arm respectively, which
puts a lower limit on the delivered resolution of the spectra based on the
tabulated values of the delivered resolutions, which is 4350, 7450, and 5300 for
the UVB, VIS and NIR-arm respectively\footnote{\url{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/inst.html}}.
The slit width also sets the width of the atmospheric sky lines and determines
the amount of light lost due to the wavelength-dependent seeing PSF extending
outside the coverage of the slit, where the width of sky-lines is always set by
the slit width whereas both the delivered resolution and the slitloss changes
for the better as the seeing PSF drops below the slit width.
For atmospheric conditions delivering a seeing PSF with a FWHM of 0\farc9
observed with a 0\farc9 slit only 76.1 percent of the light will enter the slit,
meaning that for almost all observations a slitloss correction is required. We
describe how slitlosses were corrected for in Sect. \ref{postproc}.
For accurate measurements of velocity widths, a precise instrumental resolution
is required and this becomes better when the delivered seeing is better than the
slit width. We discuss this in Sect. \ref{resolution}.

Due to mechanical issues, the atmospheric dispersion corrector (ADC) were
disabled from 1st of August 2012. Only GRB~100728B was affected by the failing
ADC prior to disablement, resulting in a lower-than-nominal throughput. To avoid
chromatic slit losses due to atmospheric dispersion, all subsequent observations
have been carried out at parallactic angle. A consequence of this is that for
all observations following 1st of August 2012, the source moves across the
spatial direction of the slit as a function of wavelength. This effect has been
modeled in the extraction procedure, as described in \ref{extract}.

We list the overview of all the observations in Tab. \ref{tab:sample_overview}.
\input{tables/burst_overview.tex} and plot the positions off all the bursts on
the night sky in Fig. \ref{fig:skymap}. The central zone of avoidance due to
high galactic extinction cutoff is visible with an apparent isotropic
distribution of the bursts away from the galactic center except in the upper
right corner, where Earth casts a shadow on the sky as seen from the telescope.

A subset of the bursts presented here  have already had their hosts investigated
in \citet{Kruhler2015}, for which extractions of the hosts exist. The focus of
the data presented here are on the afterglows themselves and in the absence of a
clear afterglow, the host. We will not, however, investigate the hosts.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data processing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In this section we describe how the final data products are produced and
subsequently post-processed. $\sim$ 30 percent of the data released here have
been reduced in a manner similar to what is described in \citet{Kruhler2015}.
The two reduction schemes have been tested and are in good agreement. All
post-processing scripts developed for this dataset are made publicly available
at \url{https://github.com/jselsing/XSGRB_reduction_scripts}, along with
instructions of use.

Before any reductions are done, the raw object images are run through the
cosmic-ray removal algorithm \citep{VanDokkum2001} implementation,
\textit{Astro-SCRAPPY} \footnote{\url{https://github.com/astropy/astroscrappy}},
where a wide clipping radius have been used around detected cosmics to ensure that edge residuals are
robustly rejected.

The basis for the reductions is the VLT/X-shooter pipeline, version
\texttt{2.7.1} or newer \citep{Goldoni2006, Modigliani2010}. The pipeline is
managed with the Reflex interface \citep{Freudling2013} and is used for
subtraction of bias level, flat-fielding, tracing of the echelle orders,
wavelength calibrations with the use of arc-line lamps, flux calibration using
spectrophotometric standards \citep{Vernet2010, Hamuy1994}, mirror flexure
compensation, initial sky-subtraction and lastly the rectification and merging
of the orders. For the initial sky-subtraction, the background has been
estimated in regions adjacent to the object trace clear of contaminating
sources. Because of the broken ADC, for some objects there is a lot of curvature
in the object trace. This means that for some bursts, the initial sky-estimate
has been made from a limited number of pixels in the spatial direction. By doing
an initial subtracting the sky on the un-rectified image we ensure that bulk of
the sky background is not redistributed by the rectification process.

@@

For a typical observation, each of the exposures in the nodding sequence has
been reduced as single observation and then subsequently combined to form a
single image. Because this strategy is employed, we can reject outliers in the
stack and weight by the average inverse variance of the background. When
weighting images where the noise in each pixel is dominated by Poisson noise it
important to estimate the background variance in a large enough region, so that
to remove the correlation between the signal and the weights. To this end, the
weight map is generated by a running median window over the variance map, where
the trace as been masked and width of the window is chosen to be wide enough for
median variance to be generated on the basis of several hundred pixels. This
weighting scheme automatically also optimally combines images of different
exposure times or images where the background is varying, which is often the
case when a burst has been observed close to twilight

Because the background is very bright and there is a high abundance of broad
sky-lines in the NIR arm, when there are no contaminating sources in the slit,
the sky has been put back on the images and they have been reduced in pairs of
two, subtracting the two from each other, keeping the WCS static. This amounts
to the regular nodding reduction, only we can reject outliers and weight by the
averaged inverse variance map.

>>>>

\abstract{In this work we present the spectra of all $\gamma$-ray bursts (GRBs)
	afterglows that have been observed at the VLT/X-shooter up to \termdate. This
	totals 137 spectroscopic observations of 118 individual bursts, of which
	redshifts have been secured for 91 \% covering a redshift range from 0.059 to
	$\sim$ 8.5. Some of these observation are late-time observations of the
	underlying host galaxy. Based on an set of selection criteria that are unbiased
	with regards to intrinsic properties of the GRBs, the follow-up effort has been
	guided by the \textit{Swift}-telescope to produce a homogeneous sample of
	afterglows for 86 spectra. We here provide a public release of all the reduced
	spectra, including continuum estimates and telluric corrections. We provide an
	assessment of the degree of completeness with respect to the parent GRB
	population, in terms of the X-ray properties of the bursts in the sample and
	find that the sample presented here is consistent with being drawn from the
	\textit{Swift} sample. For all bursts (39) for which it is possible, we provide
	a measurement of the hydrogen column density, increasing the total number of
	published HI column density measurements by $\sim$ 50 \%. We constrain the
	fraction of dark bursts to be 27\% - 42\% \todo{Preliminary} and tentatively
	confirm previous results that larger optical darkness is correlated with
	increased X-ray absorption. This dataset provides a unique possibility to study
	the ISM across cosmic time.

}

\keywords{Gamma-ray burst: general --- galaxies: high-redshift --- ISM: molecules --- dust, extinction }

@@

Gamma-Ray bursts (GRBs) are interesting as they provide the opportunity to
derive new constraints on a very wide range of topics in astrophysics. Examples
range from small scale phenomena relating to magnetars, properties of highly
relativistic jets, hyper/supernova explosions, the interstellar medium, dust
extinction curves, starburst galaxies, chemical and molecular abundances, escape
of ionizing radiation, the ionization state of the intergalactic medium,
intervening absoption systems to standard candles in cosmology
\citep[e.g.,][]{Lyons2010, Molinari2007, HjorthBloom2012, Wijers1998,
	Prochaska2009, Savaglio2006, Ghirlanda2007}. In that sense GRB research is very
multidisciplinary.

The {\it Swift} satellite \citep{Gehrels2009}, which was launched in 2004, has
made it possible to harvest much of the great potential in using GRBs as probes
of galaxies in the intergalactic medium, which was already hinted at by results
from earlier missions \citep[e.g.,][]{Ricker2004}. A crucial aspect of the great
success of the {\it Swift} mission has been the extensive ground based
follow-up observations of the afterglows and of the host galaxies of the GRBs.
This work has involved a large community of researchers both inside and outside
the official {\it Swift} team. This fruitful collaboration has been greatly
facilitated by the open data access policy of the {\it Swift} mission.

In the beginning of the "{\it Swift} era" most of the follow-up spectroscopy was
secured using low-resolution spectrographs \citep[typically
R=$\lambda/\Delta\lambda$$<$1000, e.g.][]{Fynbo2009}. Low-resolution
spectroscopy is powerful for a number of reasons, first of all is the ability to
secure information for even very faint targets. This allows the measurements of
a number of important parameters such as redshifts, spectral slopes, and
extinction. For a handful of very bright afterglows high-resolution (typically
R$>$20000) spectra were secured, and for these events much more information
about conditions inside the host galaxies were extracted
\citep[e.g.,][]{Fiore2005, Thone2007, Prochaska2007, Vreeswijk2007,
	Castro-Tirado2010}.

The X-shooter spectrograph \citep{Vernet2011} is the first of the 2nd generation
instruments at the ESO Very Large Telescope (VLT). It is a high-throughput
echelle spectrograph covering the full spectral range from the UV atmospheric
cutoff around 300 nm to the near-infrared (near-IR) K-band ending around 2500 nm
with three spectral arms (UVB, VIS, NIR). It was designed very much with
transient source follow-up in mind as the fading luminosities of such sources
makes it urgent to secure as extensive coverage as possible in the shortest
possible time. At the same time, the resolution was designed to be in the range
4000--9000 in order to be able to get a large useful spectral range between the
many sky-background emission lines in the red and near-IR spectral ranges.

In this paper, we present the results of a dedicated effort over the years 2009
-- 2017 to use the X-shooter spectrograph to secure spectroscopic observations
of afterglows and host galaxies of GRBs detected by {\it Swift}. We here make
all data resulting from the survey, publicly available in reduced form.

The paper is organized in the following way: In Sect.~\ref{sample} we describe
the sample including the sample selection criteria and the observational
strategy. In Sect.~\ref{obs}, we describe the observations and the instrumental
setups, and in Sect.~\ref{proc} we detail the methodological strategies adopted
in the data reduction process and auxiliary material. In Sect.~\ref{results} we
describe the results of the survey, i.e. the efficiency of the follow-up effort
and the characteristics of the observed bursts. We also here assess the
completeness of the finally realized sample. Finally, we offer our conclusions
in Sect.~\ref{conclusions}.

@@

Being of transient nature, it is difficult to impose strong sample selection
criteria on GRBs, without hampering the follow-up effort. Many natural follow-up
restrictions exists already, being it weather conditions, pointing restrictions
of the telescope or poorly localized bursts as reported by the alerting
telescope. To maximize the return of the follow-up campaign, we have chosen a
few selection criteria that facilitates an unbiased selection of bursts, while
at the same time allowing for a high follow-up success rate. The importance of
defining unbiased selection criteria has been highlighted previously
\citep{Jakobsson2006b, Salvaterra2012, Hjorth2012, Vergani2015, Perley2016a},
when trying to address the intrinsic underlying distribution functions such as
the redshift distribution, host metallicity distribution, or afterglow
brightness distribution. When investigating a specific distribution function, a
high degree of \textit{completeness} is desired \citep[e.g.,][]{Perley2016b}.

In defining the selection criteria, we simultaneously aim to minimize any biases
against intrinsic astrophysical conditions while at the same time maximizing the
likelihood of successful observations, hence allowing us to obtain a higher
degree of completeness. By restricting the selection criteria to conditions
local to the Milky Way and therefore independent of intrinsic GRB properties,
the aim is that the sample collected represents the underlying distribution of
GRBs in a fair way. The selection criteria used here is based on previous,
similar studies \citep{Jakobsson2006b, Fynbo2009, Hjorth2012}. We characterize
the sample completeness in Sect. \ref{completeness}.

Our selection criteria are:

\begin{enumerate}
	\item GRB triggered by \textit{Swift}.
	\item XRT started observing within 10 minutes after the GRB; an XRT position must be distributed within 12 hr.
	\item The target must be visible for at least 60 minutes at least 30 degrees above the horizon, with the Sun below -12 degrees.
	\footnote{Note that in the P84 proposal the criteria have
	been stated a bit differently, the visibility constraint being replaced by a
	declination + Sun angle constraint. The below criteria are however those
	defining the sample.}
	\item No closeby bright stars.
	\item Galactic $A_V \lesssim 0.5$ mag.
\end{enumerate}

Our ability to observe GRB afterglows is strongly dependent on the timing and
the precision of the target positions delivered by the triggering facilities. By
selecting only bursts that have been triggered on board the \textit{Swift} space
telescope \citep{Gehrels2004}, based on the Burst Alert Telescope (BAT)
\citep{Barthelmy2005}, we  start out with a homogeneously selected sample where
burst characteristics are delivered immediately, allowing for an informed
follow-up strategy. Despite the complexity of the triggering mechanism on board
\textit{Swift} \citep{Band2006, Coward2013a}, attempts at inferring properties
of the underlying GRB population based on the detection thresholds and
triggering algorithms have been made \citep{Lien2014, Graff2016}. Restricting
the follow-up effort to bursts detected by \textit{Swift}, we therefore ensure
that the limitations of the parent sample is well studied.

Because the localization accuracy of BAT is 1 - 4 arcminutes
\citep{Barthelmy2005}, a secure host identification is in general impossible
based on BAT alone. We therefore additionally require an X-ray position from the
X-ray Telescope (XRT) \citep{Burrows2005} to be distributed to the GCN network
\citep{Barthelmy2000} within 10 hours and to account for observing constraints
on \textit{Swift} that XRT began observations within 10 minutes. The additional
timing requirement of the XRT follow-up means that all bursts in our sample have
detected X-ray afterglows. Because the XRT completeness is very high
\citep{Burrows2007}, this cut should not alter the parent sample significantly.

To ensure a minimum of observability, we require that the GRB is visible from
the telescope site at Cerro Paranal, Chile, for a least 1 hour after the trigger
with the sun below $-12$ degrees. This secures time for the spectroscopic
observations to completed. Since the GRB population is isotropically distributed
on the sky as seen from Earth and because the GRB properties does not depend on
position on the sky \citep{Briggs1996}, this cut does not influence our ability
to fairly sample the underlying GRB population. The same arguments applies to
the requirement that there are no nearby bright stars. We additionally require
that the Galactic extinction is below $A_V \lesssim 0.5$ mag, based on the
extinction maps by \citet{Schlegel1998}. Because of isotropy, these additional
cuts should not influence the optical properties of the bursts themselves, only
our ability to successfully secure observations that allows us to investigate
the spectroscopic properties of GRBs.

We further discuss the effect of these selection criteria and their implication
for the compleness of the sample in Sect. \ref{completeness}.

@@

\todo{Will be written by Daniele}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{RRM observations} \label{RRM}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Under rare circumstances, the use of the rapid response mode (RRM) has been
possible. In case of a promptly visible GRB, a robotic trigger was sent to
the telescope if at the moment of the GCN notice the GRB is observable and
fulfills the following criteria.

\begin{enumerate}
	\item GRB triggered by \textit{Swift}.
	\item X-shooter mounted at the telescope
	\item GRB promptly visible
	\item Error radius < 60\arcsec.
	\item No limit on declination. 
	\item No constraint on Moon phase or distance. 
	\item Elevation in the sky > 22$^\circ$ (both now and 15 min after the trigger).
	\item Sun elevation < -12$^\circ$ (both now and 15 min after the trigger).
	\item The X-ray position must be available less than 1 hr after the GRB.
	\item The flags KNOWN\_SOURCE, COSMIC\_RAY, DEF\_NOT\_GRB are set to false (PROB\_NOT\_GRB can be true).
\end{enumerate}

The use of RRM is unique in the sense that it allowes for the detailed study of
the temporal variability of the GRB afterglow due to effect of the GRB itself on
the sorrounding medium \citep[e.g., see][for examples]{Dessauges-Zavadsky2006,
	Vreeswijk2007, Vreeswijk2013}

For 8 of the observed GRBs in the spectra released here, RRM triggered. One of
the RRM triggers is outside the statistical sample and two of the RRM trigger
are on short GRBs. The fastest response between \swift~GRB trigger and the
beginning of spectroscopic integration is for GRB~160410A, for which the delay
was only 8.4 minutes.

@@

The observations obtained for this sample have been secured with the
cross-dispersed echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on
one of the two Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal) during
the duration of this follow-up campaign. The observations have been taken during
a period of 8 years corresponding to the ESO observing periods P84 through P98
under the following programme IDs: 098.A-0055, 097.A-0036, 096.A-0079,
095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260. We
have included a few additional bursts, from the programmes 092.D-0056(A) (PI:
Rau), 092.D-0633(E) (PI: Greiner), and 095.B-0811(B) (PI: Levan). The total
collection of spectra represents \textit{all} GRBs that have been followed up by
X-shooter up to \termdate.

The first GRB followed up is GRB090313, observed on the 15th of March, 2009,
during the commissioning of X-shooter UT3. The bursts observed during the
commissioning or science verification (GRB~090313, GRB~090530, GRB~090809,
GRB~090926) are not a part of our statistical sample. The first burst observed
after science verification and X-shooter was moved to UT2, is GRB~091018, which
thereby constitute the first burst entering our statistical sample. For all
bursts that fulfill our sample selection criteria, described in Sect.
\ref{samplecrit}, spectroscopic follow-up has been attempted with X-shooter.
Various conditions can affect our ability to follow up a given burst, and a
discussion of these conditions and their consequences for the sample is
included in Sect. \ref{badbursts}.

X-shooter covers the spectral wavelength region from 300 nm to 2480 nm in a
single exposure, by separating the light into three separate spectroscopic arms
through the use of two dichroics. The ultraviolet blue (UVB) arm covers 300 -
550 nm, the visual (VIS) arm covers 550 - 1020 nm, and the near-infrared (NIR)
arm covers 1020 - 2480 nm. For some of the observations, we have applied a
$K$-band blocking filter, cutting the coverage of the NIR arm at 2100 nm. This
is done to reduce the amount of scattered background light from the thermal
infrared. For the majority of observations, a nod-and-shuffle observing scheme
has been employed, with a nodding throw of 5\arcsec. Each nodding observation
have typically been carried out in a standard ABBA pattern. For some cases,
conditions during the follow-up, either technical or weather, have necessitated
alterations to this scheme as described in App. \ref{notes}. For RRM triggers, a
slightly different observing strategy was employed. Starting as rapidly as
possible, a simple stare mode sequence was started, with 5 spectroscopic
integrations with increasing exposure times.

For the majority of the bursts, we have observed with a slit width of 1\farc0,
0\farc9, and 0\farc9 for the UVB, VIS, and NIR-arm respectively. This sets a
lower limit on the delivered resolution of the spectra based on the tabulated
values of the delivered resolutions, which is 4350, 7450, and 5300 for the UVB,
VIS and NIR-arm
respectively\footnote{\url{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/inst.html}}. %The slit width sets the width of the atmospheric sky lines and determines the amount of light lost due to the wavelength-dependent seeing PSF extending outside the coverage of the slit, where the width of sky-lines is always set by the slit width whereas both the delivered resolution and the slitloss changes for the better as the seeing PSF drops below the slit width. For atmospheric conditions delivering a seeing PSF with a FWHM of 0\farc9 observed with a 0\farc9 slit only 76.1 percent of the light will enter the slit, meaning that for almost all observations a slitloss correction is required.
We describe how slitlosses were corrected for in Sect. \ref{postproc}. For
accurate measurements involving line profiles, knowledge of the precise
instrumental resolution is required. The resolution becomes better than the
nominal one, listed above, when the delivered seeing is smaller than the
projected width of the slit on the sky. We discuss how we determine the
effective instrumental resolution in Sect. \ref{resolution}.

Due to a mechanical failure, the atmospheric dispersion corrector (ADC) was
disabled from 1st of August 2012. Only GRB~100728B was affected by the failing
ADC prior to disablement, resulting in a lower-than-nominal throughput. To avoid
chromatic slit losses due to atmospheric dispersion, all subsequent observations
have been carried out at parallactic angle. A consequence of this, is that for
all observations following 1st of August 2012, the centroid of spectrum changes
position across the spatial direction of the slit as a function of wavelength.
This effect has been modeled in the extraction procedure, as described in Sect. 
\ref{extract}.

\input{tables/burst_overview.tex}

We provide an overview of all the observations in Table
\ref{tab:sample_overview} and plot the positions of all the bursts on the
celestial sphere in galactic coordinates in Fig. \ref{fig:skymap}. Away from
central zone of avoidance due to high galactic extinction cutoff (marked in
white), the GRB positions have an apparent isotropic distribution except in the
upper left quadrant, which cannot be observed from Paranal.

A subset of the spectra presented here, primarily host observations, have
already been published in \citet{Kruhler2015}. We include independent
reductions of them here for completeness.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data processing} \label{proc}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In this section we describe how the final data products are produced and
subsequently post-processed. %$\sim$ 30 percent of the data released here have
% been reduced in a manner similar to what is described in \citet{Kruhler2015}.
% The two reduction schemes have been tested and are in good agreement.
All post-processing scripts developed for this dataset are made publicly
available at \url{https://github.com/jselsing/XSGRB_reduction_scripts}, along
with instructions of use.

Before any reductions are initiated, the raw object images are run through the
cosmic-ray removal algorithm \citep{VanDokkum2001} implementation,
\textit{Astro-SCRAPPY} \footnote{\url{https://github.com/astropy/astroscrappy}},
where a wide clipping radius have been used around detected cosmics to ensure that edge residuals are
robustly rejected.

The basis for the reductions is the VLT/X-shooter pipeline, version
\texttt{2.7.1} or newer \citep{Goldoni2006, Modigliani2010}. The pipeline is
managed with the Reflex interface \citep{Freudling2013} and is used for
subtraction of bias level, flat-fielding, tracing of the echelle orders,
wavelength calibrations with the use of arc-line lamps, flux calibration using
spectrophotometric standards \citep{Vernet2010, Hamuy1994}, mirror flexure
compensation, initial sky-subtraction and lastly the rectification and merging
of the orders. For the initial sky-subtraction, the background has been
estimated in regions adjacent to the object trace clear of contaminating
sources. Because of the broken ADC, for some objects there is a lot of curvature
in the object trace along the dispersion axis of the slit. This means that for
some bursts, the initial sky-estimate has been made from a limited number of
pixels in the spatial direction. By doing an initial subtracting the sky on the
un-rectified image we ensure that bulk of the sky background is not
redistributed by the rectification process.

@@

For a typical observation, each of the exposures in the nodding sequence has
been reduced as single observation and then subsequently combined to form a
single image. Because this strategy is employed, we can reject outliers in the
stack and weight by an averaged measure of the inverse variance of the
background. When weighting images where the noise in each pixel is dominated by
Poisson noise it important to estimate the background variance in a large enough
region, so that any correlation between the signal and the weights are removed.
To this end, the weight map is generated by a running median window over the
variance map, where the trace as been masked and width of the window is chosen
to be wide enough for median variance to be generated on the basis of several
hundred pixels. This weighting scheme automatically also optimally combines
images of different exposure times or images where the background is varying,
which is often the case when a burst has been observed close to twilight

For the NIR arm, because the background is very bright and there is a high
abundance of broad sky-lines, when there are no contaminating sources in the
slit, the sky has been put back on the images and they have been reduced in
pairs of two, subtracting the two from each other, keeping the WCS static. This
amounts to the regular nodding reduction, only we can reject outliers and weight
by the averaged inverse variance map.
